A weakly structured stem for human origins in Africa

### 2022-03-07 paper.tex
Update paper.tex
I tried to make the language about hominin, modern human, present-day human, and contemporary human more precise, less ambiguous, and more consistent.

\begin{abstract}
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135kya, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral \emph{Homo spp.} populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
\end{abstract}

@@

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}. These
approaches are more challenging to interpret and fit to data. However, new
population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with modern humans in Eurasia, several recent articles
have shown that introducing an archaic ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific archaic remains, and the possibility of
finding ancient DNA evidence (e.g., \citet{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic admixture model, leaving out other plausible models
(Figure~\ref{fig:supp-possible-models} and \citet{Henn2018-rf}). Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African ancestry (Figure \ref{fig:1}C). While ancient DNA from
Eurasia has helped us understand early human history outside of Africa, there
is no comparably ancient DNA to elucidate early history in Africa.

Here, we therefore aim to discriminate among a broader set of demographic
models by interrogating the genomes of contemporary populations. We take as our
starting point 4 classes of models (single population expansion, single
population expansion with regional persistence, archaic admixture, and
multi-regional evolution, Figure~\ref{fig:supp-possible-models}), using 290
genomes from southern, eastern, and western Africa as well as Eurasia. By
including geographically and genetically diverse populations across Africa, we
infer demographic models that explain more aspects of genetic diversity in more
populations than previously reported. These analyses confirm the inadequacy of
tree-like models and provide an opportunity to directly evaluate a wide range
of alternative models.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for archaic gene flow from Neanderthals in non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550kya from a common
stem. For each population, we computed low-order allele frequency and linkage
disequilibrium (LD) statistics that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

We began with a model of geographic expansion from a single ancestral,
unstructured source followed by migration between populations, without allowing
for contribution from an archaic African lineage or population structure prior
to the expansion (Figure~\ref{fig:supp-possible-models}A). As expected
\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
(log-likelihood ($LL$) $\approx -189,400$, Table~\ref{tab:supp-single-origin}).
We next explored a suite of models in which population structure predates the
differentiation of contemporary groups, including models allowing for ancestral
reticulation (Figure~\ref{fig:supp-possible-models}B), archaic admixture
(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
(Figure~\ref{fig:supp-possible-models}D).

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. 
We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto the more ancient
`stem' groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

@@

To account for population structure prior to 135kya, three of our four models allowed for two or more
``stem'' populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or ``merger'') events from the
other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic introgression and
fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, real uncertainty). To convey model uncertainty, we
highlight features of the two inferred models with high likelihoods. These are referred to as the
``multiple-merger'' and the the ``continuous-migration'' models. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related, and each branch contributes large amounts to all contemporary populations (Figure 3). 
Because of this 
relatedness, only $1\%$ to $4\%$ of genetic differentiation among modern populations can be traced back
to this early population structure (Figure 3 and Supp. Information)

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/fig3.pdf}
    \caption{
        \textbf{Predicted population structure over time.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains present-day structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled present-day
        populations that aligns with drift between past populations
        (here Nama and Mende, see Section~S\ref{sec:f4} for details and additional
        comparisons in Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        present-day genetic differentiation.
        Most present differentiation dates back to the last 100kya.
    }
    \label{fig:3}
\end{figure}

@@

Previous studies have found support for archaic admixture in Africa using
two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
\citep{Speidel2019-nj}. However, none of these studies considered a weakly
structured stem. We validated our inferred models with additional independent
approaches. We find that the observed cSFS (conditioned on the derived allele
being carried in the Neanderthal sample) is very well-described by the merger
model (Figures~\ref{fig:4}A-C and
\ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
even though this statistic was not used in the fit. Both models actually outperform
archaic models fit directly to the cSFS (for example, compare with Figure~1 in
\citet{Durvasula2020-td}).

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and early ancient DNA samples from Africa, could further distinguish
the archaic admixture model from the weakly-structured-stem model.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S from southern Africa moved northward thereby encountering the Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
modern groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
The second $\approx100$kya between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

Following these merger events, the stems subsequently fracture into
subpopulations which then appear to persist over the past $\sim120$kya. These
subpopulations can be linked to modern-day groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Gumuz has a
$0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
(Stem 1E) 150kya versus $0.03$ probability of being inherited from the ‘southern’
subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
(see Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}).

\subsection*{Contrasting archaic admixture and a weakly structured stem}

Evidence for archaic admixture in Eurasia has bolstered the plausibility of
archaic admixture having also occurred in Africa. For this reason, previous
work has focused on archaic admixture to explain patterns of polymorphism
inconsistent with a single origin model. Here, we have shown that
weakly-structured-stem models better capture these patterns.
They also help explain an
ecological riddle posed by the archaic admixture model. Neanderthal populations
were separated from early \emph{Homo sapiens} by thousands of kilometers and
continental geographic barriers. By contrast, an archaic hominin population in
Africa would need to have stayed in relative reproductive isolation from the
ancestral human lineage over hundreds of thousands of years despite closer geographic
proximity and reproductive compatibility. The weakly-structured-stem model
resolves this ecological riddle by accommodating continuous or recurrent
contact between two or more groups present in Africa.

There is evidence for both deleterious and adaptive archaic-derived alleles in
modern genomes in the form of a depletion of Neanderthal ancestry in regulatory
regions \citep{Petr2019-xo} or an increased frequency of archaic-related
haplotypes such as at \emph{EPAS1} among Tibetans (e.g.,
\citet{Zhang2021-xx}). Under previous
archaic African admixture models, the estimated 8--10\% introgression rate is
much higher than Neanderthal gene flow, and would have plausibly been fertile
ground for dramatic selection for or against archaic-derived haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
structured stem would have occurred continuously over much longer periods.
Patterns of polymorphism that are inconsistent with the single-stem model
predictions have been used to infer putative archaic admixed segments
\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
selection against such segments \citep{Wall2019-ao}, and pervasive positive
selection \citep{Schrider2017-kl}. However, such approaches are subject to high
false positives in the presence of population structure with migration
\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
a weakly-structured-stem model within Africa.

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also Section \ref{morpho_diff}). This correspondence allows predictions of our model to be related to the fossil record. The fossil
record of Africa is sparse during the time period of the stems, but of the
available fossils, some are very similar to present-day humans (e.g., from Omo
Kibish, Ethiopia\tdwcomment{add citations: Day (1969) doi.org/10.1038/2221135a0 and Vidal et al. (2022) doi.org/10.1038/s41586-021-04275-8}), others are similar in some features but not others (e.g.,
from Jebel Irhoud, Morocco\tdwcomment{add citations: Hublin et al. (2017) doi.org/10.1038/nature22336 and Richter et al. (2017) doi.org/10.1038/nature22335}), and others are very different (e.g., from
Dinaledi, South Africa\tdwcomment{add citations: Berger et al. (2015) doi.org/10.7554/eLife.09560 and Dirks et al. (2017) doi.org/10.7554/eLife.24231.001}) in morphology. If, as our model predicts, the
genetic differences between the stems were comparable to those among
present-day human populations, the most morphologically divergent fossils are unlikely to
represent branches that contributed appreciably to present-day human
ancestries.

>>>>

\begin{abstract}
While it is now broadly accepted that \emph{Homo sapiens} origined within Africa, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, contemporary
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135kya\tdwcomment{The standard notation for anyone who studies the past (e.g., paleoanthropologists, geologists) for thousands of years ago is ka, and Ma for millions of years ago. Given this, I think we should use these throughout the paper. I can make the changes, but I didn't want to do so without discussing them}, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral \emph{Homo spp.} populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
\end{abstract}

@@

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}. These
approaches are more challenging to interpret and fit to data. However, new
population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence (e.g., \citet{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
(Figure~\ref{fig:supp-possible-models} and \citet{Henn2018-rf}). Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African ancestry (Figure \ref{fig:1}C). While ancient DNA from
Eurasia has helped us understand early human history outside of Africa, there
is no comparably ancient DNA to elucidate early history in Africa.

Here, we therefore aim to discriminate among a broader set of demographic
models by interrogating the genomes of contemporary populations. We take as our
starting point 4 classes of models (single population expansion, single
population expansion with regional persistence, archaic hominin admixture, and
multi-regional evolution, Figure~\ref{fig:supp-possible-models}), using 290
genomes from southern, eastern, and western Africa as well as Eurasia. By
including geographically and genetically diverse populations across Africa, we
infer demographic models that explain more aspects of genetic diversity in more
populations than previously reported. These analyses confirm the inadequacy of
tree-like models and provide an opportunity to directly evaluate a wide range
of alternative models.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550kya from a common
stem. For each population, we computed low-order allele frequency and linkage
disequilibrium (LD) statistics that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

We began with a model of geographic expansion from a single ancestral,
unstructured source followed by migration between populations, without allowing
for contribution from an African archaic hominin lineage or population structure prior
to the expansion (Figure~\ref{fig:supp-possible-models}A). As expected
\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
(log-likelihood ($LL$) $\approx -189,400$, Table~\ref{tab:supp-single-origin}).
We next explored a suite of models in which population structure predates the
differentiation of contemporary groups, including models allowing for ancestral
reticulation (Figure~\ref{fig:supp-possible-models}B), archaic hominin admixture
(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
(Figure~\ref{fig:supp-possible-models}D).

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. 
We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), contemporary populations cannot be easily mapped onto the more ancient
`stem' groups as only a small proportion of drift between contemporary populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

@@

To account for population structure prior to 135kya, three of our four models allowed for two or more
``stem'' populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or ``merger'') events from the
other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic hominin introgression and
fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, real uncertainty). To convey model uncertainty, we
highlight features of the two inferred models with high likelihoods. These are referred to as the
``multiple-merger'' and the the ``continuous-migration'' models. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic hominin admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related, and each branch contributes large amounts to all contemporary populations (Figure 3). 
Because of this 
relatedness, only $1\%$ to $4\%$ of genetic differentiation among contemporary populations can be traced back
to this early population structure (Figure 3 and Supp. Information)

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/fig3.pdf}
    \caption{
        \textbf{Predicted population structure over time.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see Section~S\ref{sec:f4} for details and additional
        comparisons in Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last 100kya.
    }
    \label{fig:3}
\end{figure}

@@

Previous studies have found support for archaic hominin admixture in Africa using
two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
\citep{Speidel2019-nj}. However, none of these studies considered a weakly
structured stem. We validated our inferred models with additional independent
approaches. We find that the observed cSFS (conditioned on the derived allele
being carried in the Neanderthal sample) is very well-described by the merger
model (Figures~\ref{fig:4}A-C and
\ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
even though this statistic was not used in the fit. Both models actually outperform
archaic hominin admixture models fit directly to the cSFS (for example, compare with Figure~1 in
\citet{Durvasula2020-td}).

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic hominin admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
hominin admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and early ancient DNA samples from Africa, could further distinguish
the archaic hominin admixture model from the weakly-structured-stem model.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S from southern Africa moved northward thereby encountering the Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
contemporary groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
The second $\approx100$kya between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

Following these merger events, the stems subsequently fracture into
subpopulations which then appear to persist over the past $\sim120$kya. These
subpopulations can be linked to contemporary groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Gumuz has a
$0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
(Stem 1E) 150kya versus $0.03$ probability of being inherited from the ‘southern’
subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
(see Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}).

\subsection*{Contrasting archaic hominin admixture and a weakly structured stem}

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of
archaic hominin admixture having also occurred in Africa. For this reason, previous
work has focused on archaic hominin admixture to explain patterns of polymorphism
inconsistent with a single origin model. Here, we have shown that
weakly-structured-stem models better capture these patterns.
They also help explain an
ecological riddle posed by the archaic hominin admixture model. Neanderthal populations
were separated from early \emph{Homo sapiens} by thousands of kilometers and
continental geographic barriers. By contrast, an archaic hominin population in
Africa would need to have stayed in relative reproductive isolation from the
ancestral human lineage over hundreds of thousands of years despite closer geographic
proximity and reproductive compatibility. The weakly-structured-stem model
resolves this ecological riddle by accommodating continuous or recurrent
contact between two or more groups present in Africa.

There is evidence for both deleterious and adaptive archaic-hominin-derived alleles in
contemporary genomes in the form of a depletion of Neanderthal ancestry in regulatory
regions \citep{Petr2019-xo} or an increased frequency of archaic-hominin-related
haplotypes such as at \emph{EPAS1} among Tibetans (e.g.,
\citet{Zhang2021-xx}). Under previous
African archaic hominin admixture models, the estimated 8--10\% introgression rate is
much higher than Neanderthal gene flow, and would have plausibly been fertile
ground for dramatic selection for or against archaic-hominin-derived haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
structured stem would have occurred continuously over much longer periods.
Patterns of polymorphism that are inconsistent with the single-stem model
predictions have been used to infer putative archaic admixed segments
\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
selection against such segments \citep{Wall2019-ao}, and pervasive positive
selection \citep{Schrider2017-kl}. However, such approaches are subject to high
false positives in the presence of population structure with migration
\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
a weakly-structured-stem model within Africa.

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also Section \ref{morpho_diff}). This correspondence allows predictions of our model to be related to the fossil record. The fossil
record of Africa is sparse during the time period of the stems, but of the
available fossils, some are very similar to contemporary humans (e.g., from Omo
Kibish, Ethiopia\tdwcomment{add citations: Day (1969) doi.org/10.1038/2221135a0 and Vidal et al. (2022) doi.org/10.1038/s41586-021-04275-8}), others are similar in some features but not others (e.g.,
from Jebel Irhoud, Morocco\tdwcomment{add citations: Hublin et al. (2017) doi.org/10.1038/nature22336 and Richter et al. (2017) doi.org/10.1038/nature22335}), and others are very different (e.g., from
Dinaledi, South Africa\tdwcomment{add citations: Berger et al. (2015) doi.org/10.7554/eLife.09560 and Dirks et al. (2017) doi.org/10.7554/eLife.24231.001}) in morphology. If, as our model predicts, the
genetic differences between the stems were comparable to those among
contemporary human populations, the most morphologically divergent fossils are unlikely to
represent branches that contributed appreciably to contemporary human
ancestries.
